---
title: "Toronto Street Tree Population"
subtitle: "Evaluating the health of City of Toronto's street trees by taxonomy and size"
date: "August 6, 2026"
date-format: long
author: 
  name: "Jocelyn Cao, Harry Parmar"
format:
  revealjs:
    pdf-separate-fragments: true
    theme: solarized
    css: styles.css
    highlight-style: github
    slide-number: true
bibliography: ref.bib
nocite: |
  @*
---
```{r}
library(tinytable)
library(here)
library(dplyr)
library(broom)
```

## Background

- Harry

- City of Toronto Canopy goal

## 10-20-30 Rule

- Harry

## DBH Distributions

- Harry

## Research Question

- Harry

## Data Source

- Harry

## Methods

- Data cleaning

- Exploratory data analysis

- Linear regression model

## Data Cleaning

- Cleaned variable names using janitor package [@R-janitor]

- Created a new table with only necessary variables

  - Object ID: unique ID of tree
  
  - Ward: tree location by geographical district
  
  - Botanical name: formal scientific name of tree
  
  - DBH: diameter of tree's trunk; indicative of tree's maturity
  
## Data Cleaning (Continued)

- Derived new variables based on existing variables:

  - Region based on ward (e.g., Wards 6, 8, 15, 16, 17, and 18 are considered North York)
  
  - Species, genus, and family based on botanical name using classification by WorldFlora package [@WorldFlora2020]

## Exploratory Data Analysis

- Proportion of majority **species** in each ward

- Must be under **0.10** to pass 10-20-30 rule

```{r}
ward_tab <- readRDS(here("models", "ward_table.rds"))
ward_tab@data |> 
  select(ward, region, max_species_prop, dominant_species) |> 
  tt() |>
    style_tt(fontsize = 0.7)
```

## Exploratory Data Analysis (pt.2)

- Proportion of majority **genus** in each ward

- Must be under **0.20** to pass 10-20-30 rule

```{r}
ward_tab <- readRDS(here("models", "ward_table.rds"))
ward_tab@data |> 
  select(ward, region, max_genus_prop, dominant_genus) |> 
  tt() |>
    style_tt(fontsize = 0.7)
```

## Exploratory Data Analysis (pt.3)

- Proportion of majority **family** in each ward

- Must be under **0.30** to pass 10-20-30 rule

```{r}
ward_tab <- readRDS(here("models", "ward_table.rds"))
ward_tab@data |> 
  select(ward, region, max_family_prop, dominant_family) |> 
  tt() |>
    style_tt(fontsize = 0.7)
```


## Linear Regression Model {.smaller}

- Is there a relationship between the excess percentage of majority tree type (species, genus, family) and mean ward trunk diameter (dbh)?

```r{}
model_2_excess <- lm(mean_dbh ~ species_excess + genus_excess + family_excess, 
  data = ward_metrics_refined)
```

```{r}
model_2 <- readRDS(here("models", "model_2.rds"))
summary(model_2)
```

## Linear Regression Model (pt.2)

- R-squared: 

- p-value:

## Linear Regression Model (pt.3)

```{r}
model_2 <- readRDS(here("models", "model_2.rds"))
tidy(model_2) |> 
  tt() |> 
  style_tt(fontsize = 0.75)
```

- p-value must be less than 0.05 to be statistically significant

- for every 1 percentage point a ward exceeds the **20% genus threshold**, the average trunk diameter **increases by 0.90 cm**

- for every 1 percentage point a ward exceeds the **30% family threshold**, the average trunk diameter **decreases by ~2.03 cm**

## Linear Regression Model (pt.4)
```{r}
plot_4 <- readRDS(here("models", "plot_4.rds"))
plot_4
```
## Findings
- Wards that severely breach the 20% genus threshold rely heavily on mature, larger trees

- This reflects historical urban forestry practices, such as over-planting of Acer (Maples), that built large street canopy footprints through a  single genus.

## Findings
- Exceeding the 30% family threshold is associated with significantly smaller average trunk diameters.




## Summary

## References
::: {#refs style="font-size: 0.6em;"}
:::

